---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# latexr

The goal of latexr is to translate LaTeX formulas to R code. (Formerly known as `latex2r`; the main function is still `latex2r()`.)

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("rodrigoesborges/latexr")
```

This is a very young package so it may not work as expected if you try to translate
things outside of the supported syntax. Please refer to [Supported LaTeX](#supported-latex) 
section for more information about it.

## Examples

Just some basic funcionality: translate LaTeX to R.
```{r example}
library(latexr)
latex2r("\\beta_1^{\\frac{x+1}{x^2 \\cdot y}}")
```

With a combination of `parse()` and `eval()` you can evaluate the translated expression.
Of course, names must be bound to a value if we expect this to work.
```{r example2}
eval(parse(text = latex2r("\\pi * \\sin(\\frac{x}{2})")), envir = list(x = pi))
```

There is an extra feature which is possible due to R is so permissive. `latex2fun()` 
receives a LaTeX expression that represents the definition of a mathematical function
and returns an R function that computes the function value and has arguments representing
all the variables involved in the function.

```{r example3, dpi=500, out.width="75%", fig.align="center"}
x = seq(-2*pi, 2*pi, length.out = 500)
f = latex2fun("\\sin{a * x}^2 + \\cos{b * x} ^2")
print(f)

y = f(x = x, a = 2, b = 3)
plot(x, y, type = "l")
```

This is experimental but I think it is so cool that it is worth a chance in the package. 
For those who like to play with R most weird features, they would find
the [source code](R/latex2fun.R) is a nice place.

In addition, if you call `latex2r(interactive=TRUE)` it launches a REPL that you can
use interactively.

## Supported LaTeX

Only a small subset of LaTeX expressions are suported so far. However,
these are enough to define a very wide set of mathematical functions. 

### Greek letters supported

The following greek letters are supported as identifiers (variable names).

```{r}
latexr:::get_pkg_data('GREEK_KEYWORDS')
```

### Syntax supported

You can use the following operators

* Unary:
  + `+` and `-`.
* Binary:
  + `+`, `-`, `*`, `/`, `^`, and `_`.
* Grouping:
  + `{...}`, `\left{...\right}`, `(...)`, and `\left(...\right)`.
* Absolute value:
  + `\left|x\right|`.

And the following functions

* Unary:
  + `\sqrt`, `\log`, `\ln`, `\sin`, `\cos`, `\tan`, `\cosh`, `\sinh`, and `\tanh`.
  + Statistical: `\bar{x}` and `\overline{x}` (mean), and `\tilde{x}` (median).
  
* Binary:
  + `\frac{...}{...}`, `\cdot`, and `\times`.
  + Rolling sum: `\sum_{k}^{}{x}` translates to `data.table::frollsum(x, k)`.

Spacing commands (`\;`, `\,`, `\:`, `\quad`, `\qquad`) are ignored.

#### Notes

* The operator `_` is used to represent subscripts. While you can do $5_2$ in LaTeX, it 
is not allowed in the package since a subscript on a number does not make sense.  
Only variable names (such as `x` or `\\pi`) can have subscripts.
* In latex you can write `\sqrt[p]{x}` to represent the p-th root. However this is 
not allowed in this package (at least for now). To represent a p-th root you can use `x^{1/p}`.

### Some remarks

#### Implicit multiplication

**tl;dr:** `xy` is understood as `x` times `y`.

A previous version of this package required multiplication to be explicit. 
For example, `xy` would have been understood as an identifier called `xy`. 
Now, all identifiers, except from special ones (greek letters), are of one character 
only. If you do `abc^5` it will be understood as `a*b*c^5`.

In addition, you can still pass an explicit multiplication operator such as `*`,
`\times` or `\cdot`.

#### Explicit grouping

Although something like `\sin5` renders as $\sin5$ and we all understand this means
sine of 5, we require explicit grouping with `{}` or `()` to avoid ambiguity in the function
argument. What if I write `\sin5a`? Does it mean a times the sine of 5 or the sine
of 5 times a? Explicit grouping is a simple solution to eliminate this ambiguity.

#### Special treatment for some characters

Since the numbers <img src="https://render.githubusercontent.com/render/math?math=\pi"> 
and <img src="https://render.githubusercontent.com/render/math?math=e"> are so common 
in mathematical expressions they are treated as constant numbers and not as names of variables.

In `R` <img src="https://render.githubusercontent.com/render/math?math=\pi"> is a 
built-in constant number and <img src="https://render.githubusercontent.com/render/math?math=e">
is obtained with `exp(1)`. 
See the next example

```{r}
latex2r("\\sin{2 * \\pi * t}")
latex2r("e * x")
latex2r("e^{x + i * y}")
```

But note that complex numbers are not supported (yet?).

#### Logarithm of different base

If you write `\\log(x)` it will be interpreted as the natural logarithm of `x`.
If you write `\\log_n(x)` it will be interpreted as the logarithm of `x` with base `n`.
The alias `\\ln(x)` also works.

```{r}
latex2r("\\log(x + 1)")
latex2r("\\log_2(x + 1)")
latex2r("\\ln(x + 1)")
```

#### Statistical notation

`\\bar{x}` and `\\overline{x}` translate to `mean(x)`, and `\\tilde{x}`
translates to `median(x)`. Since real data often contains missing values,
both accept `na.rm = TRUE`:

```{r}
latex2r("\\bar{x} + \\tilde{y}")
latex2r("\\bar{x} + \\tilde{y}", na.rm = TRUE)
```

For windowed calculations, `\\sum_{k}^{}{x}` translates to
`data.table::frollsum(x, k)`, a rolling sum of window `k`.

```{r}
latex2r("\\sum_{3}^{}{x}")
```

#### Unicode input (MathQuill and friends)

Visual formula editors such as MathQuill emit raw Unicode characters instead
of LaTeX commands. `latex2r()` normalizes them automatically before scanning:
`−`, `–` and `—` become `-`; `×` becomes `\times`; `⋅` and `·` become `\cdot`;
`÷` becomes `/`; Greek letters become the corresponding commands; and accented
vowels such as `ā`, `ã` and `â` become `\bar{a}`, `\tilde{a}` and `\hat{a}`
(other accents fall back to the base letter).

```{r}
latex2r("2 × 3")
latex2r("α + β")
normalize_mathquill("ā")
```

The `normalize_mathquill()` function is exported so the mapping can be
inspected and reused.

#### Inspecting the parsed expression

`latex2ast()` runs the same pipeline as `latex2r()` but returns the parsed
abstract syntax tree instead of the R translation, which is useful for
debugging formulas.

```{r}
latex2ast("\\bar{x}")
```

### Examples

```{r echo=FALSE, eval=FALSE}
examples = c("x + y", "\\sin(x) + \\cos(y)", "\\sin(x)^2 + \\cos(y)^2", 
             "\\sqrt{2x\\pi}", "\\log(z)", "\\log_a(\\frac{x^5}{y})", 
             "\\frac{1}{\\sigma\\sqrt{2\\pi}}e^{\\frac{(x - \\mu)^2}{2\\sigma^2}}",
             "\\beta_1^{\\frac{x+1}{x^2 \\cdot y}}")
examples_r = unname(vapply(examples, latex2r, character(1)))
data = data.frame(
  `LaTeX` = examples, `R Code` = examples_r
)
knitr::kable(data)
```

|LaTeX                                                          |R Code                                                           |
|:--------------------------------------------------------------|:----------------------------------------------------------------|
|`x + y`                                                        |`x + y`                                                          |
|`\sin(x) + \cos(y)`                                            |`sin(x) + cos(y)`                                                |
|`\sin(x)^2 + \cos(y)^2 `                                       |`sin(x)^2 + cos(y)^2 `                                           |
|`\sqrt{2x\pi} `                                                |`sqrt(2 * x * pi) `                                              |
|`\log(z)`                                                      |`log(z)`                                                         |
|`\ln(z)`                                                       |`log(z)`                                                         |
|`\log_a(\frac{x^5}{y})`                                        |`log((x^5) / y, base = a) `                                      |
|`\frac{1}{\sigma\sqrt{2\pi}}e^{\frac{(x - \mu)^2}{2\sigma^2}}` |`1 / (sigma * sqrt(2 * pi)) * exp(((x - mu)^2) / (2 * sigma^2))` |
|`\beta_1^{\frac{x+1}{x^2 \cdot y}}`                            |`beta_1^((x + 1) / (x^2 * y))`                                   |
|`\bar{x}`                                                      |`mean(x)`                                                        |
|`\tilde{x}`                                                    |`median(x)`                                                      |
|`\left|x + y\right|`                                           |`abs(x + y)`                                                     |
|`\sum_{3}^{}{x}`                                               |`data.table::frollsum(x, 3)`                                     |
